A* with Perfect Heuristic/Potentials in Road Networks

### 2020-07-30 paper/ch_potentials.tex
foo

\caption{Example A* search space with CH-Potentials TODO: what parametrization?}
\label{img:search-space}
\end{figure}

@@

@@

A Contraction Hierarchy (CH) is a two phase technique to efficiently compute optimal, shortest paths.
For a detailed exposition, we refer the interested reader to \cite{gssv-erlrn-12,dsw-cch-15}.
In this section, we give a high-level overview.

A CH places nodes into levels.
No edge must connect two nodes within one level.
Levels are ordered by ``importance''.
The intuition is that dead-ends are unimportant and low while highway bridges are very important and high.
An edge goes \emph{up} when it goes from a node in a lower level to higher level.
\emph{Down} edges are defined analogously.
An \emph{up-down path} is a path where only one node $v$ is more important than its neighbors.
$v$ is called the \emph{mid} node.
An \emph{up path} is a path where the last node is the mid node.
Similarily, the first node is the mid node of a \emph{down path}.
Every up and down path is an up-down path.
%
In the preprocessing phase, a CH adds \emph{shortcut} edges to the input graph $G$ to obtain $G^+$.
This is done by repeatedly contracting unimportant nodes. 
See \cite{gssv-erlrn-12} for the details.
The trick is that in $G^+$ for pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path with the same length as a shortest path in $G$.
See Figure~\ref{fig:ch} for an illustration.
We may therefore restrict our search to up-down paths in $G^+$.
This search is bidirectional.
The forward search starts from $s$ and only follows up-edges.
Similarily, the backward search starts at $t$ and only follows down-edges in reversed direction.
The two searches meet in the mid node.
Pseudo-code for the backward search, i.e., the path from $m$ to $t$, is presented in Algorithm~\ref{algo:ch-backward}.
The forward search works analogously.
%
A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
On road networks, this is the case as demonstrated in \cite{gssv-erlrn-12,dgrw-gpnc-11,dgpw-crprn-13,dsw-cch-15,hs-gbpo-18}.

@@

PHAST \cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node.
First the preprocessing phase is executed analogously to the original CH.
The query phase is divided into two steps.
The first step is analogously to the CH query:
From the target node $t$, all reachable nodes via reversed down-edges are explored.
Algorithm~\ref{algo:ch-backward} shows this first step.
The second step iterates over all CH level from top to bottom.
In each iteration, all up-edges starting within the current level are followed in reverse.
After all level are processed, the distances from all nodes towards $t$ is computed.
PHAST is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
PHAST's main advantage is that is easily be parallelized.
However, we will not consider parallelization in this paper.
We refer to \cite{dgnw-phast-13} for an in-depth experimental performance analysis.
Pseudo-code is provided in Algorithm~\ref{algo:phast}.

Using PHAST, we can compute optimal A* potentials in another way.
In the query phase, we first use PHAST to compute the distances from every node to $t$ with respect to the query edge weights $w_\ell$ and store the result in an array $H$.
In the next step, we run $A^*$ and implement the potential function as lookup in the array $H$.

This setup also works.
The potential evaluation and by extension the $A^*$ search is indeed fast.
However, the PHAST step before the search is comparatively expensive.
The reason for this is that the distance from all nodes towards $t$ are computed.
Ideally, we only want to compute the distances from the nodes explored in the $A^*$ search towards $t$.

@@

Fortunately, the PHAST computation can be done lazily using memoization as depicted in Algorithm~\ref{algo:pot}.

In a first step, we run the backward CH search from $t$ to obtain an array $B$.
$B[v]$ is the minimum down $vt$-path distance or $+\infty$, if there is no such path.
$B$ is computed as shown in Algorithm~\ref{algo:ch-backward}.

@@

Potential evaluation is an expensive operation that we want to avoid.
A potential evaluation is always paired with a queue push operation.
By avoiding push operations, we can therefore avoid potential evaluations.
This is the objective of this and the next section.

@@

%\begin{figure}

%\begin{center}
%\includegraphics{push-deg3}
%\end{center}

%\caption{$x$ is removed from the queue. $y_i$, $z$, $a_i$, and $\alpha_i$ are never part of the queue. $\beta$ and $b$ are pushed into the queue.}
%\label{fig:push-deg3}

%\end{figure}

@@

If $d(z)\neq 3$ or $z$ is in the queue, our algorithm proceeds as in the previous section.
Otherwise, there exist up to two degree chains $z,a_1,\ldots,a_p,b$ and $z,\alpha_1,\ldots,\alpha_q,\beta$ such that $a_1\neq y_k \neq \alpha_1$.
Our algorithm iteratively tries to reduce all $D[a_i]$ and $D[\alpha_i]$.
If it reaches $\beta$, $\beta$'s weight in the queue is set to $D[\beta]+\pi(\beta)$.
Analogously, if $b$ is reached, $b$'s weight is set to $D[b]+\pi(b)$.
If $b$ respectively $\beta$ are not reached, our algorithm does nothing.
%The situation is depicted in Figure~\ref{fig:push-deg3}.

\subsection{Skip Nodes Outside of Largest Biconnected Component}

@@

\section{Extensions}
\label{sec:extensions}

@@

\subsection{Forbidden Turns}
\label{sec:no-turns}

@@

\subsection{Predicted Traffic}
\label{sec:predicted-traffic}

>>>>

\caption{Example A* search space with CH-Potentials}
\label{img:search-space}
\end{figure}

@@

In this section, we first describe Contraction Hierarchies, then PHAST, and finally CH-Potentials.

@@

\subsection{Contraction Hierarchy (CH)}

A CH is a two phase technique to efficiently compute optimal, shortest paths.
It is not flexibile, i.e., $w_q=w_\ell$.
For a details, we refer the interested reader to \cite{gssv-erlrn-12,dsw-cch-15}.
In this section, we give an overview.

A CH places nodes into levels.
No edge must connect two nodes within one level.
Levels are ordered by ``importance''.
The intuition is that dead-ends are unimportant and at the bottom while highway bridges are very important and at the top.
An edge goes \emph{up} when it goes from a node in a lower level to higher level.
\emph{Down} edges are defined analogously.
An \emph{up-down path} is a path where only one node $v$ is more important than its neighbors.
$v$ is called the \emph{mid} node.
An \emph{up path} is a path where the last node is the mid node.
Similarily, the first node is the mid node of a \emph{down path}.
Every up and down path is an up-down path.
%
In the preprocessing phase, a CH adds \emph{shortcut} edges to the input graph $G$ to obtain $G^+$.
This is done by repeatedly contracting unimportant nodes and adding shortcuts between its neighbors.
See~\cite{gssv-erlrn-12} for the details.
The trick is that in $G^+$ for pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path with the same length as a shortest path in $G$.
See Figure~\ref{fig:ch} for a proof scetch.
From every shortest path (red) in $G$, a up-down path of equal length in $G+$ (blue) can be constructed.
We may therefore restrict our search to up-down paths in $G^+$.
This search is bidirectional.
The forward search starts from $s$ and only follows up-edges.
Similarily, the backward search starts at $t$ and only follows down-edges in reversed direction.
The two searches meet in the mid node.
Pseudo-code for the backward search, i.e., the path from $m$ to $t$, is presented in Algorithm~\ref{algo:ch-backward}.
The forward search works analogously.
%
A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
On road networks, this is the case as demonstrated in \cite{gssv-erlrn-12,dgrw-gpnc-11,dgpw-crprn-13,dsw-cch-15,hs-gbpo-18}.

@@

PHAST \cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node.
First the preprocessing phase is executed analogously to the original CH.
The query phase is divided into two steps.
The first step is analogously to the CH query:
From the target node $t$, all reachable nodes via reversed down-edges are explored.
Algorithm~\ref{algo:ch-backward} shows this first step.
The second step iterates over all CH level from top to bottom.
In each iteration, all up-edges starting within the current level are followed in reverse.
After all level are processed, the distances from all nodes towards $t$ is computed.
PHAST is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
PHAST's main advantage is that is easily be parallelized.
However, we will not consider parallelization in this paper.
We refer to \cite{dgnw-phast-13} for an in-depth experimental performance analysis.
Pseudo-code is provided in Algorithm~\ref{algo:phast}.
Using PHAST, we can compute optimal A* potentials in another way.
In the query phase, we first use PHAST to compute the distances from every node to $t$ with respect to the query edge weights $w_\ell$ and store the result in an array $H$.
In the next step, we run $A^*$ and implement the potential function as lookup in the array $H$.

This PHAST-based algorithm works.
The potential evaluation and by extension the $A^*$ search is indeed fast.
However, the PHAST step before the search is comparatively expensive.
The reason for this is that the distance from all nodes towards $t$ are computed.
Ideally, we only want to compute the distances from the nodes explored in the $A^*$ search towards $t$.

@@

Fortunately, the PHAST computation can be done lazily using memoization as depicted in Algorithm~\ref{algo:pot}.
In a first step, we run the backward CH search from $t$ to obtain an array $B$.
$B[v]$ is the minimum down $vt$-path distance or $+\infty$, if there is no such path.
$B$ is computed as shown in Algorithm~\ref{algo:ch-backward}.

@@

Potential evaluation is an expensive operation that we want to avoid.
A potential evaluation is always paired with a queue push operation.
By avoiding push operations, we can therefore avoid potential evaluations.
This is the objective of this section.

\subsection{Skip Degree Two Nodes}

@@

@@

If $d(z)\neq 3$ or $z$ is in the queue, our algorithm proceeds as in the previous section.
Otherwise, there exist up to two degree chains $z,a_1,\ldots,a_p,b$ and $z,\alpha_1,\ldots,\alpha_q,\beta$ such that $a_1\neq y_k \neq \alpha_1$.
Our algorithm iteratively tries to reduce all $D[a_i]$ and $D[\alpha_i]$.
If it reaches $\beta$, $\beta$'s weight in the queue is set to $D[\beta]+\pi(\beta)$.
Analogously, if $b$ is reached, $b$'s weight is set to $D[b]+\pi(b)$.
If $b$ respectively $\beta$ are not reached, our algorithm does nothing.

\subsection{Stay in Largest Biconnected Component}

@@

\section{Algorithm Extensions}
\label{sec:extensions}

@@

\subsection{Forbidden Turns and Turn Costs}
\label{sec:no-turns}

@@

\subsection{Predicted Traffic or Time-Dependent Routing}
\label{sec:predicted-traffic}
